---
title: "My report"
author: "Faustin GASHAKAMBA"
date: "12/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reproducible Research Project 1

### Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

### Loading and preprocessing the data

Loading the necessary libraries and then read the data into a variable called "Activitiy"

```{r initialization, echo=TRUE}
library(ggplot2)
library(plyr)
library(lubridate)
activity <- read.csv("activity.csv")
```

Processing the data by formatting the dates appropriately and deal with NA values

```{r cleaning, echo=TRUE}
activity$date <- as_date(activity$date)
clean <- activity[!is.na(activity$steps),]
```

### Question 1: What is mean total number of steps taken per day?

Calculate the total number of steps taken per day

```{r Summarize-data, echo=TRUE}
## summarizing total steps per date
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")
```

Make a histogram of the total number of steps taken each day

```{r makeGraph-1, echo=TRUE}
## Creating the historgram of total steps per day
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```
Calculate and report the mean and median of the total number of steps taken per day

```{r Mean, echo=TRUE}
## Mean of Steps
as.integer(mean(sumTable$Steps))
```
The average number of steps taken each day was 10766 steps.

```{r Median, echo=TRUE}
## Median of Steps
as.integer(median(sumTable$Steps))
```
The median number of steps taken each day was 10765 steps.

### Question 2: What is the average daily activity pattern?

Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r TimeSeries, echo=TRUE}
library(plyr)
library(ggplot2)
##pulling data without nas
clean <- activity[!is.na(activity$steps),]

##create average number of steps per interval
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))

##Create line plot of average number of steps per interval
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r Number-of-steps, echo=TRUE}
##Maximum steps by interval
maxSteps <- max(intervalTable$Avg)
maxSteps
```

The maximum number of steps for a 5-minute interval is 206 steps.

```{r Max-of-steps, echo=TRUE}
##Which interval contains the maximum average number of steps
intervalTable[intervalTable$Avg==maxSteps,1]
```
The 5-minute interval which had the maximum number of steps is 835
